\documentclass[landscape]{article}
\usepackage{fullpage}
\pagestyle{empty}

\newcommand{\optionalmath}[1]{\ifmmode#1\else$#1$\fi}
\let\mathge\ge
\let\mathle\le
\def\ge{\optionalmath\mathge}
\def\gt{\optionalmath>}
\def\le{\optionalmath\mathle}
\def\lt{\optionalmath<}
\def\C{\optionalmath\clubsuit}
\def\D{\optionalmath\diamondsuit}
\def\H{\optionalmath\heartsuit}
\def\S{\optionalmath\spadesuit}
\def\NT{\ifmmode\mathsc{nt}\else\textsc{nt}\fi}
\def\HCP{\textsc{hcp}}
\def\DP{\textsc{dp}}
\def\f{\optionalmath\star}
\def\gf{\optionalmath\copyright}
\def\s{\textsc{s}}

\begin{document}

\title{Bridge from the ground up}
\author{Stephen Hicks}

\begin{minipage}[t]{0.33\columnwidth}
\section{Scoring (Rubber Bridge)}
\subsection{contracts}
only points bid go below the line\\
\C \D -- 20 per trick (``minors'')\\
\H \S -- 30 per trick (``majors'')\\
\NT\kern3.5pt -- 40 for first, 30 thereafter

\subsection{slam bonus}
small slam (12 tricks): 500/750 if vulnerable\\
grand slam (13 tricks): 1000/1500

\subsection{overtricks}
normally worth 20/30 depending on suit.\\
if doubled, redoubled, or vulnerable, each\\
\phantom{m}trick is worth 100/200/400, regardless of suit.

\subsection{undertricks}
not doubled -- 50/100 if vulnerable\\
doubled -- 100, 200, 200, 300, \ldots $\quad$ if not vul.,\\
\phantom{doubled -- }200, 300, \ldots\phantom{,100, 200} $\quad$ if vulnerable\\
redoubled -- 2x doubled

\subsection{rubbers}
game = 100 points below line\\
two games = rubber\\
bonus -- 700 if 2--0,\\
\phantom{bonus -- }500 if 2--1\\
vulnerable after winning a game

\end{minipage}
\vrule\hskip .005\columnwidth
\begin{minipage}[t]{0.33\columnwidth}
\section{The Auction}

dealer begins with first bid.  bidding proceeds clockwise
around table.  possible bids are a number followed by a suit
(\C,\D,\H,\S,\NT), double, redouble, or pass.  bidding on
the number of tricks past six (``book'') for the partnership
to take with their chosen trump suit.

\subsection{suit order}
1\C, 1\D, 1\H, 1\S, 1\NT,\\
2\C, 2\D, 2\H, 2\S, 2\NT,\\
\phantom{mmnnnl}\kern1pt\ldots,\\
7\C, 7\D, 7\H, 7\S, 7\NT

\subsection{game bids}
the bids 5\C, 5\D, 4\H, 4\S, and 3\NT\ are known as
game bids because they are worth at least 100 points.  lower
bids are known as part-scores.  game bids are to be desired
if at all possible.

\subsection{high card points (HCP)}
A--4, K--3, Q--2, J--1\\
all aces -- +1 bonus\\
no aces -- -1 penalty\\

distribution points (\DP):\\
if opening, void--2, singleton--1, doubleton--0\\
if responding \textbf{w/ fit}, void--5, singl.--3, doubl.--1\\
only 3-card fit -- -1 penalty (at minimum)\\
no \DP\ for \NT\ contracts.

\end{minipage}
\vrule\hskip .005\columnwidth
\begin{minipage}[t]{0.33\columnwidth}
\section{Gameplay}
member of partnership who first opens a suit is known as the
declarer.  player left of declarer leads first (any card is
allowed, including trump).  after first lead, declarer's partner
turns up his hand as dummy for declarer to play from.

\section{Basic Bidding}
13\HCP\ to open, 7\HCP\ to respond\\

partner who knows more about other hand is captain and ultimately
decides where to play (by asking for more information).\\

forcing bids (denoted \f) require a response from partner unless
opponents interfere.\\
%game-forcing bids (denoted \gf) require a contract at least at the game level.\\

games in major or \NT\ typically require 26\HCP.\\
games in a minor suit typically require 29\HCP.\\
small slams typically require 33\HCP.\\
grand slams typically require 37\HCP.\\

open with longest suit or \NT\ if balanced.\\
strong 2-level opening with 23\HCP\\

respond with fit, suggested new suit (need \ge10\HCP\ for 2 level),
or 1\NT \f\ if no other possible response (\lt10\HCP).\\
jump-raise (game-forcing) for fit with 13\HCP.\\
captain may invite game (one less than game bid) to ask partner
for 2--3 more points than previously shown.
\end{minipage}

\vskip 1pc
This document is available online at 
\verb!http://www.physics.cornell.edu/~shicks/bridge.pdf!

\eject

\begin{minipage}[t]{0.33\columnwidth}
\section{Goren Standard Bidding}
\subsection{opening}
5-card majors required to open.\\
\begin{tabular}{rl}
  1\C\D & 13--22\HCP, \ge3 cards\\
  1\H\S & 13--22\HCP, \ge5 cards\\
  1\NT & 16--18\HCP, balanced\\
  \f\ 2\C & \ge23\HCP, any shape, forcing\\
  2\D\H\S & 6--12\HCP, \ge6 cards, no voids\\
  2\NT & 21--22\HCP, balanced\\
  3\C\D\H\S & \lt 10\HCP, \ge7 cards\\
\end{tabular}
\subsection{responses}
to a suit, if fit is discovered:\\
\begin{tabular}{rl}
  jump 1 level & 7--10\HCP\\
  jump 2 levels & 13--16\HCP\\
  jump to 4-level & \le9\HCP, 5-card support\\
  jump shift & \ge19\HCP\\
\end{tabular}\\

to a suit, with no fit:\\
\begin{tabular}{rl}
  new suit at 1-level & \ge6\HCP\\
  new suit at 2-level & \ge10\HCP\\
  1\NT & 6--10\HCP\\
  2\NT & 13--15\HCP, balanced\\
  3\NT & 16--18\HCP, balanced\\
\end{tabular}\\

to 1\NT:\\
\begin{tabular}{rl}
  2\NT & 7--9\HCP, balanced?\\
  3\NT & 10--14\HCP, balanced\\
  4\NT & \ge 15\HCP (slam?)\\
  \f\ 2\C & Stayman (4-card major)\\
  \f\ 2\D\H & Jacoby Transfer to \H\S (\ge 5 cards)\\
  3\C\D\H\S & 10\HCP, long suit\\
  4\H\S & \le 10\HCP, long suit\\
\end{tabular}\\

to 2\NT:\\
\begin{tabular}{rl}
  3\NT & 4--8\HCP\\
  4\NT & \ge 9\HCP\ (slam?)\\
\end{tabular}\\
or similar Stayman/Jacoby at 3-level.
\end{minipage}
\begin{minipage}[t]{.33\columnwidth}
\subsection{strong 2\C}
response:\\
\begin{tabular}{rl}
  2\D & weak, \le4\HCP\\
  2\H\S,3\C\D & \ge5\HCP, \ge5 card suit\\
  3\NT & 5--8\HCP, no long suit\\
  2\NT & \ge9\HCP, balanced, slam?\\
\end{tabular}

opener's rebids after 2\C:\\
\begin{tabular}{rl}
  raise to game & found fit, minimal hand\\
  raise below game & found fit, slam?\\  
\end{tabular}\\

may want to arrange for transfer, rather than direct suit
bid.

\subsection{Stayman 2\C}
2\C\ after 1\NT\ is Stayman, asking for 4-card major\\
\begin{tabular}{rl}
  2\D & no 4-card major\\
  2\H & 4-card \H\ suit\\
  2\S & 4-card \S\ suit, \lt 4 \H's\\
\end{tabular}

\subsection{Jacoby transfer 2\D\H}
2\D\ or 2\H\ after 1\NT\ is Jacoby transfer, so that\\
stronger hand is declarer.\\
\begin{tabular}{rl}
  2\D & bid 2\H\\
  2\H & bid 2\S\\
\end{tabular}\\
responder will continue with normal responses afterwards.

\subsection{competitive bidding}
takeout double\f\ if just enough to open after opponent bids.
partner names longest and strongest suit or 1\NT\ if balanced.
partner may only pass if confident in penalty double.\\

overcall only if good, long, strong suit (\ge 10\HCP\ at
1-level, \ge 13\HCP at 2-level).\\

%redouble after T/O shows strong hand, but a suit bid is more descriptive.

\end{minipage}\hskip .01\columnwidth
\begin{minipage}[t]{.33\columnwidth}

\subsection{penalty doubles}
count tricks.  only double if can set by two.\\
\begin{tabular}{cc}
opening & defensive tricks\\
1\C\D\H\S & 3\\
1\NT & 4\\
takeout dbl & 3\\
overcall & 1\\
raise of opener & 1\\
preempt & 0\\
\end{tabular}\\
beware of doubling slam bids

\subsection{cuebids}
after suit agreed upon, bids in other suits show
control (void, ace) and slam interest.

\subsection{blackwood/gerber}
4\NT\ (or 4\C) asks for aces in a suit (\NT) contract.
responses follow:\\
\begin{tabular}{rl}
  5\C (4\D) & 0 or 4 aces\\
  5\D (4\H) & 1 ace\\
  5\H (4\S) & 2 aces\\
  5\S (4\NT) & 3 aces\\
\end{tabular}\\
asker may then bid 5\NT (5\C) to ask for kings. any other
bid is terminal.

\subsection{Michael's cuebid}
indicates a two-suited (at least 5-5) hand by bidding same
suit immediately after opponent.  cuebid of minor indicates
both majors.  cuebid of major indicates other major and
unspecified minor (responder may bid 2\NT\ to ask for minor).\\

only use Michael's cuebid with \le 9\HCP\ or \ge 16\HCP.
prefer overcall with 10--16\HCP.

\end{minipage}

\end{document}

