\begin{sheet}{Competition}\setcounter{section}0

{\strut\hfill
\begin{ptable}\headings{5-8|9-12|13-17|18+}
\row{balanced}
  \bid(13-17,lgreen){Takeout double}
  \bid*(18,lred){Takeout double}\\
\row{5 card suit}
  \bid*(9-17,lblue){Direct overcall}
  \bid(18,lred){Takeout double}\\
\row{6+ card suit}
  \bid(5-8,lorange){Jump overcall}
  \bid(9-17,lblue){Direct overcall}
  \bid*(18,lred){Takeout double}\\
\row{Two-suited}
  \bid(9-12,lpurple){Michael's/Unusual}
  \bid*(13-17,lblue){}
  \bid(18,lyellow){Michael's/Unusual}\\
\end{ptable}\hfill\strut}

\begin{columns}{2}
\begin{column}
\section{Direct Overcalls}
Direct suit overcalls are nonforcing, showing 10--17 points and a 5-card suit.

\section{Takeout Doubles}
A double over a below-game suit bid (through 4\d) when partner has never made a
non-passing bid is for takeout.  Partner \emph{must respond} unless opponent intervenes$^*$.
\strut\vskip-1.3pc\strut
\begin{description}\itemsep-4pt
\item[0--5]pass if opp intervenes
\item[0--9]bid longest suit
\item[10--12]jump in longest suit
\item[13+]cue-bid
\end{description}
\strut\vskip-1.3pc\noindent
With a stopper and no unbid 4-card major:
\strut\vskip-1.3pc\strut
\begin{description}\itemsep-4pt
\item[6--10]1\nt with stopper
\item[10--12]2\nt with stopper
\item[13--16]3\nt with stopper
\item[10--15]game-forcing
\item[16--17]slam-invitational
\item[18+]slam-forcing
\end{description}

{\noindent
\begin{ptable}\headings{0-5|6-9|10-12|13-16|17+}
\row{4 card}
  \bid(0-5,lorange){\pass$^*$}
  \bid(6-9,lgreen){bid suit}
  \bid(10-12,lyellow){jump suit}
  \bid(13-17,lred){cue-bid}\\
\row{balanced}
  \bid(6-9,lpurple){1\nt}
  \bid(10-12,lorange){2\nt}
  \bid(13-16,lblue){3\nt}
  \bid*(17,lred){}\\
\end{ptable}}
\vskip8pt\noindent
(With 0--9 points, responder may double-jump with a 6-card major.)
\end{column}

\begin{column}
\section{Two-suited bids}
With a two-suited hand and either strong or weak strength (9--12 or 17+),
bid Michael's cuebid or Unusual 2\nt.  Partner should assume the weak
range until the overcaller shows otherwise with a voluntary bid.  Both
bids are forcing.

\subsection{Michael's cuebid}
Over a minor opening, cuebidding the opener's suit shows 5--5 in the majors.
Over a major opening, cuebidding the opener's suit shows 5 in the other major
and 5 in an undisclosed minor.  Partner may bid \nt to ask which minor.

\subsection{Unusual 2\nt}
Unusual 2\nt\ shows a 5--5 hand in the two lower unbid suits.
\end{column}
\end{columns}

\vskip-8pt

\end{sheet}
